---
title: "Stats_Modelling_2"
author: "Lorenz Wolf"
date: "3/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

```{r}
getwd()
```
# Data Exploration
```{r}
# get quick overview
head(dat)
summary(dat$bmi)
summary(dat$rest_pulse) 
summary(dat$stimulated_pulse)
```
```{r}
# summary of coffee drinkers
summary(subset(dat, dat$coffee_reg==1))
# summary of non-coffee drinkers
summary(subset(dat, dat$coffee_reg==0))
```

```{r}
pulse_bmi <- ggplot(dat, aes(x=bmi, y=rest_pulse)) + 
 			geom_point()
```
```{r}
pulse_bmi <- ggplot(dat, aes(x=bmi, y=rest_pulse, colour='rest pulse')) + 
 			geom_point() + xlab('BMI') + ylab('Pulse')

plot = pulse_bmi + geom_point(data=dat, aes(x=bmi, y=stimulated_pulse, colour='stimulated pulse')) + ggtitle('Pulses vs BMI')
plot
```
First note, that heart rates and bmi's are all in a reasonable range, nothing obviously strange.
We can observe that generally the stimulated pulse is higher than the rest pulse. It also looks like there is a correlation between stimulated pulse and bmi while this does not seem to be the case for the rest pulse.

```{r}
rest_coffee <- ggplot(dat, aes(x=bmi, y=rest_pulse, colour=coffee_reg)) + 
 			geom_point() + xlab('BMI') + ylab('Rest Pulse')+theme(legend.position = "none")

stim_coffee <- ggplot(dat, aes(x=bmi, y=stimulated_pulse, colour=coffee_reg)) + 
 			geom_point() + xlab('BMI') + ylab('Stimulated Pulse')

figure <- ggarrange(rest_coffee,stim_coffee 
                    , ncol = 2, nrow = 1)
annotate_figure(figure, top = 'Pulses for coffee and non-coffee drinkers')
```
```{r}
box_bmi <- ggplot(dat, aes(x=coffee_reg, y=bmi, fill=coffee_reg)) + geom_boxplot() + xlab('Coffee')+ ylab('bmi') + ggtitle('BMI for coffee and non-coffee drinkers')
box_bmi
```
```{r}
pulse_diff <- ggplot(dat, aes(x=bmi, y=stimulated_pulse - rest_pulse, colour=coffee_reg)) + 
 			geom_point() + xlab('BMI') + ylab('Pulse Difference')+ theme(legend.position = "none")
box_diff <- ggplot(dat, aes(x=coffee_reg, y=stimulated_pulse-rest_pulse, fill=coffee_reg)) + geom_boxplot()+ xlab('Coffee')+ ylab('Pulse Difference')
figure <- ggarrange(pulse_diff,box_diff
                    , ncol = 2, nrow = 1)
figure
```

Looks like non-coffee drinkers have larger bmi. Need to be careful with this in the analysis. For example suppose the stimulant has a larger effect on non-coffee drinkers, the this could lead to the wrong conclusion that the stimulant has larger effect on people with higher bmi. Hence probably want to fit a model with coffe_reg as covariate and see what happens. Note however, this could also be the other way around, but intuitively not the case - would need clinicians input (i.e. can receptors get used to stimulation through coffee consume? INTUITIVELY YES -based on own experience:) ).

```{r}
# number of coffee drinkers and non-coffee drinkers
length(subset(dat$coffee_reg, dat$coffee_reg==1))
length(subset(dat$coffee_reg, dat$coffee_reg==0))
```
Coffee consumers dominate the study. How does this affect the study?? Could this lead to underestimation of the stimulant's effect?? 

```{r}
# boxplot of both pulses for coffee vs non-coffee
box_rest <- ggplot(dat, aes(x=coffee_reg, y=rest_pulse, fill=coffee_reg)) + geom_boxplot()+ xlab('Coffee')+ ylab('Rest Pulse') + theme(legend.position = "none")
box_stim <- ggplot(dat, aes(x=coffee_reg, y=stimulated_pulse, fill=coffee_reg)) + geom_boxplot() + xlab('Coffee')+ ylab('Stimulated Pulse')

figure <- ggarrange(box_rest, box_stim
                    , ncol = 2, nrow = 1)
annotate_figure(figure, top = 'Pulses for coffee and non-coffee drinkers')

```
Checking mean and variance before and after stimulation for coffe and non coffee drinkers can observe that more drastic change for non-coffee drinkers.

# Fitting Models
## Model 1
```{r}
# fitting model
model1 <- lm(stimulated_pulse~rest_pulse+bmi   ,data=dat)
```
```{r}
# checking summary and residual plots of model
summary(model1)
plot(model1)
```
```{r}
AIC(model1)
```

## Manual IWLS
```{r}
# feeding in data
x <- dat$bmi
y <- dat$stimulated_pulse-dat$rest_pulse
X <- cbind(1,x)
```

```{r}
# function for deviance
D <- function(p){
  2* sum(y/p - log(y/p) -1)
}

# function to compute convergence criterion from old and new deviance
convergence <- function(o,n){
  abs(n-o)/(abs(n)+0.1)
}

# least squares estimator as initial guess
eta <- 1/(dat$stimulated_pulse-dat$rest_pulse) 
beta_start <- solve(t(X)%*%X)%*%t(X)%*%eta
beta <- beta_start #initial guess

# initial deviance
oldD <- D(y)
conv_crit <- 1
# IWLS
while(conv_crit>10^(-8)){
  
 eta <- X%*%beta #estimated linear predictor
 mu <- 1/eta #estimated mean response
 z <- 2*eta - y * eta^2 #form the adjusted variate
 w <- (mu^2) #weights
 lmod <- lm(z~x, weights=w) #regress z on x with weights w
 beta <- as.numeric(lmod$coeff) #new beta
 print(beta) #print out the beta estimate every iteration 
 newD <- D(1/(X%*%beta)) #compute new deviance
 conv_crit <- convergence(oldD, newD) #compute convergence criterion

 oldD <- newD #update old deviance
 }
```
```{r}
# estimate deviance and obtain standard errors forparameterestimates
phi_hat <- oldD/(length(dat$bmi)-2) #estimate of dispersion
cov_beta <- phi_hat * solve( t(X) %*% diag(as.vector(w))%*%X) #covariance matrix of beta
beta.sd <- sqrt(as.vector(diag(cov_beta))) #standard errors of beta

```

```{r}
# manually compute 95% confidence intervals
lower <- beta-1.96*beta.sd
upper <- beta+1.96*beta.sd

cat('95% confidence interval for beta_1 is (',format(lower[1]),',',format(upper[1]),')\n')
cat('95% confidence interval for beta_2 is (',format(lower[2]),',',format(upper[2]),')\n')

```

## Fitting Model 2 suggested by statistician
```{r}
model2 <- glm(stimulated_pulse-rest_pulse~bmi,Gamma(link = "inverse"), data=dat)
summary(model2)
plot(model2)
```
```{r}
# plotting model fit
bmi_p <- seq(15.9,25,0.5)
pred_pulse_diff <- predict(model2, list(bmi=bmi_p), type='response')
plot(dat$bmi,dat$stimulated_pulse-dat$rest_pulse)
lines(bmi_p, pred_pulse_diff, col='red')
```
## Fitting own models

### GLM adding coffee as predictor
```{r}
# add coffee as predictor
model3 <- glm(stimulated_pulse-rest_pulse~bmi+coffee_reg,Gamma(link = "inverse"), data=dat)
summary(model3)
plot(model3)
```
```{r}
# compute blow up point of model
cat('Model 3 for non-coffee drinkers blows up at a bmi of',-model3$coefficients[1]/model3$coefficients[2], '\n')
cat('Model 3 blows up for coffee drinkers at a bmi of',-(model3$coefficients[1]+model3$coefficients[3])/model3$coefficients[2])
```

```{r}
# plot predictions of model 3 for appropriate bmi.
# predictions for coffee drinkers
bmi_coffee <- seq(15.9,25,0.1)
pred_coffee <- predict(model3, newdata = data.frame('bmi'=bmi_coffee,'coffee_reg'=as.factor(1)), type='response')
df_coffee <- data.frame('bmi'=bmi_coffee, 'predicted'=pred_coffee)
# predictions for non-coffee drinkers
bmi_NOcoffee <- seq(19.7,24,0.1)
pred_NOcoffee <- predict(model3, newdata = data.frame('bmi'=bmi_NOcoffee,'coffee_reg'=as.factor(0)), type='response')
df_NOcoffee <- data.frame('bmi'=bmi_NOcoffee, 'predicted'=pred_NOcoffee)
# scatter plot and predictions
model3_plot <- ggplot(dat, aes(x=bmi, y=stimulated_pulse - rest_pulse, colour=coffee_reg)) + 
 			geom_point() + xlab('BMI') + ylab('Pulse Difference')
model3_plot <- model3_plot + geom_line(data=df_coffee, aes(x=bmi,y=predicted, colour='1'))+
  geom_line(data=df_NOcoffee, aes(x=bmi,y=predicted, colour='0')) + ggtitle('Fit of Model 3')
model3_plot
```
```{r}
# obtain confidence intervals for parameters
confint(model3)
```
```{r}
model3$coefficients
```

GLM with gamma distribution and identity link. This leads to a higher residual deviance and larger AIC as model3.
```{r}
model4 <- glm(stimulated_pulse-rest_pulse~bmi+coffee_reg,Gamma(link = "identity"), data=dat)
summary(model4)
plot(model4)
```

```{r}
# plot predictions of model 3 for appropriate bmi.
# predictions for coffee drinkers
pred_coffee <- predict(model4, newdata = data.frame('bmi'=bmi_coffee,'coffee_reg'=as.factor(1)), type='response')
df_coffee <- data.frame('bmi'=bmi_coffee, 'predicted'=pred_coffee)
# predictions for non-coffee drinkers
pred_NOcoffee <- predict(model4, newdata = data.frame('bmi'=bmi_NOcoffee,'coffee_reg'=as.factor(0)), type='response')
df_NOcoffee <- data.frame('bmi'=bmi_NOcoffee, 'predicted'=pred_NOcoffee)
# scatter plot and predictions
model4_plot <- ggplot(dat, aes(x=bmi, y=stimulated_pulse - rest_pulse, colour=coffee_reg)) + 
 			geom_point() + xlab('BMI') + ylab('Pulse Difference')
model4_plot <- model4_plot + geom_line(data=df_coffee, aes(x=bmi,y=predicted, colour='1'))+
  geom_line(data=df_NOcoffee, aes(x=bmi,y=predicted, colour='0')) + ggtitle('Fit of Model 4')
model4_plot
```
Even though the linear link is easier to interpret, it does not


GLM with gamma distribution and log link. This leads to a higher residual deviance and larger AIC as model3.
```{r}
# add coffee as predictor
model5 <- glm(stimulated_pulse-rest_pulse~bmi+coffee_reg,Gamma(link = "log"), data=dat)
summary(model5)
plot(model5)
```
### Analyse adding terms (coffee, interaction term)
```{r}
myglm <- glm(stimulated_pulse-rest_pulse~., family=Gamma(link = "inverse"), data=dat )
summary(myglm)
```
```{r}
anova(myglm,test='Chisq')
```
```{r}
drop1(myglm, test='Chisq')
```
Both predictors appear to be significant. Now check if adding interaction term leads to a better model.
```{r}
add1(myglm,~.^2, test='Chisq')
```
Interaction term between coffee and bmi not significant.

### Consider other normal linear models
```{r}
mylm <- lm(stimulated_pulse-rest_pulse~bmi, data=dat)
summary(mylm)
plot(mylm)
```
```{r}
# try box cox transform
library(MASS)
bc <- boxcox(dat$stimulated_pulse-dat$rest_pulse ~ dat$bmi, lambda=seq(0,0.4,0.01))
# extract max lambda value
bc$x[which.max(bc$y)]
# [1] 0.2222222
x <- 0.2222222
bcmodel <- lm(((stimulated_pulse-rest_pulse)**x - 1)/x ~ bmi, data=dat)
summary(bcmodel)
plot(bcmodel)
```
0 just about not in 95% confidence interval.
QQ plot and residuals look better than for initial linear model. Small Adjusted R-squared. Insignificant Intercept term.
```{r}
mylm2 <- lm(stimulated_pulse~rest_pulse+bmi+coffee_reg, data=dat)
summary(mylm2)
plot(mylm2)
```





